---
title: "Perturbation Table Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
```

Sidebar {.sidebar}
=====================================

### Parameter Input

```{r}


textInput("label", "Parturbation Label", paste("CKM_D",3,"_V",1.05*100,sep=""))

sliderInput("D", label = "Maximum Perturbation (=D)",
            min = 1, max = 10, value = 3, step = 1)
numericInput(inputId="V", label="Variance (=V)", 
             value=1.05, min = 0.1, max = 5, step = 0.1, width="100px")
sliderInput("js", label = "Treshold (=js)",
            min = 0, max = 5, value = 1, step = 1)

numericInput(inputId="pstay", label="Preset probability of frequencies not to be perturbed (=pstay)",              value=NA, min = 0.05, max = 0.99, step = 0.05, width="200px")
#checkboxInput("pstayset", "Preset probability of frequencies not to be perturbed (=pstay)", value=FALSE, width = NULL)
#test <-  reactive(input$pstayset)
#if (reactive(test()) == TRUE) pstay <- NA
#sliderInput("pstay", label = "Set probability (=pstay)",
#            min = 0.05, max = 0.95, value = 0.5, step = 0.05)

checkboxInput("mono", "Set Monotony", value = TRUE, width = NULL)
radioButtons("optim", "Set of objective functions (=optim)",
               c("Standard" = 1,
                 "Alternative" = 4))

# TODO: input for pTableSize with respect to ncat.

params <- reactive({
  pt_create_pParams(D=input$D, V=input$V, js=input$js, pstay=input$pstay, label=input$label, mono=input$mono, optim=as.integer(input$optim))
})
res <- reactive( {
  pt_create_pTable(params = params())
})
```



**Save the Output**

```{r download}
downloadHandler(
    filename = function() {
      paste("graph-", Sys.Date(), ".pdf", sep="")
    },
    contentType = "image/pdf",
    outputArgs = "Graph",
    content = function(file) {
      fifi_plot(res(), file=file)
      # TODO: replace by new argument 'file'
    }
  )
cat("\n")
downloadHandler(
    filename = function() {
      paste("ptable-", Sys.Date(),".csv", sep="")
    },
    contentType = "text/csv",
    outputArgs = "pTable",
    content = function(file) {
      out <- res()
      #write.csv(out@pTable[i>=0,c('i','j','p','v','p_int_ub'),] ,file, row.names = FALSE, quote=FALSE)
      write.table(format(out@pTable[i>=0,c('i','j','p','v','p_int_ub'),], digits=8), file, sep=";", dec=".", row.names = FALSE, col.names = TRUE, quote=FALSE)
      # TODO: replace by pt_export
      #pt_export(input = out, file = file)
    }
  )
```


Results {.tabset}
=====================================

Columns {.tabset}
-------------------------------------



### Perturbation Table 

```{r}
library(DT)
renderTable(
  res()@pTable[i>=0,c("i","j","p","v","p_int_lb","p_int_ub"),] , digits = 8)#,
#options = list(scrollX = TRUE, sScrollY = '75vh', scrollCollapse = TRUE))
```

### Constraints Check {data-width=100}

```{r}
renderTable({
  res()@empResults[i>0,,]
}, digits=3)
```

### Input Object {data-width=175}

```{r}
renderPrint({
  params()
})
```

### Legend {data-width=75}

parameter     | description
---------------- | ------------------------
i  | original frequency
j | target frequency
p | perturbation probability
v | perturbation value / noise
p_int_lb | lower bound of p (within each frequency i)
p_int_ub | upper bound of p (within each frequency i)
p_mean   | mean of probabilities for given original frequency
p_var | variance of probabilities for given original frequency
p_sum | sum of probabilities for given original frequency
p_stay | probability for a given original frequency not to be perturbed


Visualization {.tabset}
=====================================

Columns {.tabset}
-------------------------------------


### Transition Matrix

```{r}
renderPlot({
  fifi_plot(res(), type="t")
})
```

### Distribution

```{r}
renderPlot({
  fifi_plot(res())
})
```

### Perturbation Panel

```{r}
renderPlot({
  fifi_plot(res(), type="p")
})
```
